---
title: "Climate Exercise"
author: "Kristin Liu"
output: github_document
---



# Unit I: Climate Change Module

## Warm-up exercise: Examining CO2 trends in R

- Example from <http://climate.nasa.gov/vital-signs/carbon-dioxide/>
- Raw data from <https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt>

```{r message=FALSE}
library(tidyverse)
library(dplyr)
library(zoo)
```

```{r message=FALSE}

co2 <- 
read_table("https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt", 
                  comment="#",
                  col_names = c("year", "month", "decimal_date", "average",
                                "interpolated", "trend", "days"),
                  na = c("-1", "-99.99"))
co2
```


```{r}
ggplot(co2, aes(x = decimal_date, y = average)) + geom_line() 
```

The months the CO2 values are at the maximum are in the winter months. Trees are dormant in the winter, which causes the levels of CO2 to rise because there are no trees actively photosynthesizing to convert CO2 into oxygen. The months the CO2 values are at the maximum are in the summer, which is when trees are actively photozynthesizing to convert CO2 into O2.

The rolling average that is used in computing the trend line is the average of the maximum and minimum CO2 values for each year. The overall positive trend indicates that the rolling average has been increasing since 1960, and that the maximum and minimum CO2 values of each year have both been increasing.

------------------------

# Exercise I: Temperature Data

Each of the last years has consecutively set new records on global climate.  In this section we will analyze global mean temperature data.

Data from: <http://climate.nasa.gov/vital-signs/global-temperature>

## Question 1:

In the global temperature change data set, the columns for each data provide the year, the original mean values of temperature change in global surface temperatures (in Celcius), and lowess values. Lowess values are a smoother set of values created from a time series regression of global surface temperature changes that have been contaminated with noise.

The measurements are made by record keeping through NASA's meteorological stations and ERSST v5, which measures ocean temperatures. The associated measurement uncertainty arises from systemic biases, which can be attributed to land cover changes or technology shifts. There is also inherent measurement uncertainty due to the averaging method and GISTEMP interpolation, as well as changes in spatial coverage of the station record. 

The method used for measurement is detailed and accurate for data resolution. The temperature is measured in small intervals of a few seconds through meteorological stations, and a 5 minute average is taken from those intervals. The 5 minute averages are aggregated into hourly averages.

There is no missing data in the dataset. If data is missing, the significance of the values should be considered as well as the impact they would have on the existing data. One or two missing points of data may not affect the overall trend, unless they are extreme outliers. On the other hand, many points of missing data would leave a significant gap in the story of global temperature change. This would lead to people questioning the credibility of the dataset and accuracy of the data.

## Question 2:

Construct the necessary R code to import and prepare for manipulation the following data set: <http://climate.nasa.gov/system/internal_resources/details/original/647_Global_Temperature_Data_File.txt>

```{r}
globaltemp <- 
read_table("https://climate.nasa.gov/system/internal_resources/details/original/647_Global_Temperature_Data_File.txt",
           skip = 5, comment = "#", col_names = c("year", "no_smoothing", "lowess"), na = "NA")
globaltemp
```


## Question 3:

```{r}
ggplot(globaltemp, aes(x = year, y = lowess)) + 
  geom_line() + 
  geom_point(size = 0.3) + 
  theme_minimal(base_size = 10)
```

The overall trend of the global mean temperature has been increasing from 1880 to 2010, with only some small occasional drops in temperature. Starting at around 1920, the global mean temperature has been increasing at a consistent greater rate. This implies that recent human activity has contributed to the emission of greenhouse gases, which has led to rapid global warming and the acceleration of global temperature increase.

## Question 4: Evaluating the evidence for a "Pause" in warming?

The [2013 IPCC Report](https://www.ipcc.ch/pdf/assessment-report/ar5/wg1/WG1AR5_SummaryVolume_FINAL.pdf) included a tentative observation of a "much smaller increasing trend" in global mean temperatures since 1998 than was observed previously.  This led to much discussion in the media about the existence of a "Pause" or "Hiatus" in global warming rates, as well as much research looking into where the extra heat could have gone.  (Examples discussing this question include articles in [The Guardian](http://www.theguardian.com/environment/2015/jun/04/global-warming-hasnt-paused-study-finds), [BBC News](http://www.bbc.com/news/science-environment-28870988), and [Wikipedia](https://en.wikipedia.org/wiki/Global_warming_hiatus)). 

The IPCC observed that there was a pause in average global temperature increase between 1998 to 2012. They suggested that there was only an increase of 0.05 degrees C per decade in those 15 years compared to the 0.12 degree C increase from 1951 to 2012. There are many compelling arguments and theories that have been published in an attempt to explain this pause; some explanations include ocean upwellings that have lead to cooler ocean temperatures or natural rise-and-fall temperature cycles. However, the time period of the pause described by the IPCC does not match up with dip in global temperature change from NASA's data set. The time period between 1951 and 2012 sees a rather sharp increase in global temperatures, with some areas of small increase or leveling out of temperature. The differences between IPCC's claim and the graph from NASA's global average temperature data may be due to measurement uncertainties. 

To better help refine this argument, additional analyses from NASA on their measurement uncertainties would be helpful in determining the accuracy of the graph in order to re-compare the data with IPCC's claims. Additional data sources on global temperature between 1998 to 2012 and temperature change analysis from the IPCC would also be helpful in determining if the data supports the claim of a pause in global warming rates.

## Question 5: Rolling averages

The 5 year average is the average of the temperature over the past 5 years, while annual average is the average temperature over the course of a year. 

```{r}
globaltemp5 = globaltemp %>%
  mutate(period = cut(year, seq(1880, 2020, 5), na.pad = TRUE)) %>%
  group_by(period) %>% 
  summarise(avg = mean(lowess)) %>% 
  ggplot(aes(period, avg)) + 
  geom_point() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
globaltemp5

globaltemp10 = globaltemp %>%
  mutate(period = cut(year, seq(1880, 2020, 10), na.pad = TRUE)) %>%
  group_by(period) %>% 
  summarise(avg = mean(lowess)) %>% 
  ggplot(aes(period, avg)) + 
  geom_point() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
globaltemp10

globaltemp20 = globaltemp %>%
  mutate(period = cut(year, seq(1880, 2020, 20), na.pad = TRUE)) %>%
  group_by(period) %>% 
  summarise(avg = mean(lowess)) %>% 
  ggplot(aes(period, avg)) + 
  geom_point() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
globaltemp20
```

In all 3 graphs, the general trends show an increase of global temperature over the century as well as a steep increase in temperatures beginning in the 1970s.

# Exercise II: Melting Ice Sheets?

- Data description: <http://climate.nasa.gov/vital-signs/land-ice/>
- Raw data file: <http://climate.nasa.gov/system/internal_resources/details/original/499_GRN_ANT_mass_changes.csv>

## Question 1:

The columns of the data set are year.decimal (the year and what fraction of the year has passed), Greenland (Greenland mass measured in gigatonne), and Antarctica (Antarctica mass measured in  gigatonne). The numbers come from NASA's GRACE and GRACE Follow-On satellites. 

The uncertainty in measurement is liable to change because the data includes the latest data as well as continuously updated data as more information is collected. The resolution for the y axis is year and the resolution for the x-axis is mass in gigatonnes. The interpretation of the missing values indicates human error or an error with the measuring equipment.

## Question 2:

```{r}
ice_sheets <- 
read_csv("499_GRN_ANT_mass_changes.csv", comment = "#", 
         col_names = c("Year.decimal", "Greenland", "Antarctica"), skip = 1, na = "NA")
ice_sheets
```

## Question 3:

```{r}
ggplot(ice_sheets, aes(x = Year.decimal, Mass)) + 
  geom_line(aes(y = Greenland, color = "Greenland mass")) + 
  geom_line(aes(y = Antarctica, color = "Antarctica mass"))
```


# Exercise III: Rising Sea Levels?

- Data description: <http://climate.nasa.gov/vital-signs/sea-level/>
- Raw data file: <http://climate.nasa.gov/system/internal_resources/details/original/121_Global_Sea_Level_Data_File.txt>


## Question 1:

The data set contains the columns: altimeter type 0=dual-frequency  999=single frequency (ie Poseidon-1), merged file cycle # , year+fraction of year (mid-cycle), number of observations, number of weighted observations, GMSL (Global Isostatic Adjustment (GIA) not applied) variation (mm) with respect to TOPEX collinear mean reference, standard deviation of GMSL (GIA not applied) variation estimate (mm), smoothed (60-day Gaussian type filter) GMSL (GIA not applied) variation (mm), MSL (Global Isostatic Adjustment (GIA) applied) variation (mm) with respect to TOPEX collinear mean reference, standard deviation of GMSL (GIA applied) variation estimate (mm), smoothed (60-day Gaussian type filter) GMSL (GIA applied) variation (mm), and smoothed (60-day Gaussian type filter) GMSL (GIA applied) variation (mm) (annual and semi-annual signal removed).

The data comes from the Integrated Multi-Mission Ocean Altimeter Data for Climate Research (http://podaac.jpl.nasa.gov/dataset/MERGED_TP_J1_OSTM_OST_ALL_V2). The data set consists of sea surface heights combined with common terrestrial reference frames, which is then placed onto a geo-referenced orbit.

There is more uncertainty in this data set due to the time taken in processing the data. The resolution for the x axis is year and the resolution for the y axis is GMSL (global isostatic adjustment) variation (mm) with respect to TOPEX colinear mean reference. There is no missing data, but if there were missing values they could be interpreted as human error, measurement equipment error, or inability to collect data for a particular timepoint.

## Question 2:

```{r}
sea_levels <- read_table("http://climate.nasa.gov/system/internal_resources/details/original/121_Global_Sea_Level_Data_File.txt", skip = 46, col_names = c("altimeter type (0=dual, 999=single)","merged file cycle number", "year", "number of observations", "number of weighted observations", "GMSL", "Standard deviation of GMSL", "GMSL (GIA applied)", "standard deviation of GMSL (GIA applied)", "Smoothed GMSL GIA applied variation", "Smoothed GMSL GIA applied variation (annual, semiannual signal removed)"), comment="#")
sea_levels
```

## Question 3:

```{r}
ggplot(sea_levels, aes(x = year, y = GMSL)) + 
  geom_line() + 
  geom_point(size = 0.3) + 
  theme_minimal(base_size = 10) 
```

The graph shows that the global mean sea level been increasing between 1993 to 2016. There are slight fluctuations in sea level within each year, likely due to seasonal changes. The steady positive trend implies that global sea levels are rising as a result of global warming.

# Exercise IV: Arctic Sea Ice?

- <http://nsidc.org/data/G02135>
- <ftp://sidads.colorado.edu/DATASETS/NOAA/G02135/north/daily/data/N_seaice_extent_daily_v3.0.csv>


## Question 1:

The data set contains the columns: year, month, day, dissing, extent, and source data. Missing shows whether or not the data is absent with either a 0 (no) or 1 (yes). Extent refers to sea ice extent, which is the area of ocean where there is at least some sea ice measured in 10^6 sq km. Source data shows the source for where all the data was gathered.

The data comes from the National Snow and Ice Data Center (https://nsidc.org/data/G02135). Changes in Arctic and Antarctic sea ice can be searched through the sea ice index which provides the latest sea ice extent and data values.

There is not much uncertainty in this data set because the data is produced consistently, which makes the index time-series usable in terms of looking at long-term trends. Daily and monthly measurements are made, but monthly measurements are more reliable for long-term analyses because daily measurements may vary due to short term weather effects. The resolution of the x axis is the date and the resolution of y axis is the extent (in 10^6 sq km). There are no missing values, but is there were they would likely be due to human error or measurement error. 

## Question 2:

```{r}
sea_ice <- read_csv("N_seaice_extent_daily_v3.0.csv", col_names = c("Year", "Month", 
                                                                    "Day", "Extent", 
                                                                    "Missing", "Source Data"), 
                    skip = 2, na = "NA")
sea_ice <- sea_ice %>% 
  mutate(Date = as.Date(paste(Year, Month, Day, sep = "-"))) %>%
  select(Date, Year, Month, Day, Extent, Missing)
sea_ice
```

## Question 3:

```{r}
sea_ice_filter <- sea_ice %>%
  filter(Month == "01", Day == "01")
ggplot(sea_ice_filter, aes(x = Date, y = Extent)) + 
  geom_line() + 
  geom_point(size = 0.3) + 
  theme_minimal(base_size = 10) 
```

There has been an overall decrease in sea ice extent from 1978 to present day. Annual fluctuations are due to seasonal changes, as logically there would be less sea ice in the summer and more sea ice in the winter. However, there is a negative linear association between the years and sea ice extent. From this trend, we can infer that sea ice is melting as a result of global warming.
